---
title: "Natural events impact between 1950 and 2011"
documentclass: article
output:
  pdf_document: default
  html_document: default
classoption: a3paper
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE);
knitr::opts_chunk$set(cache = TRUE);
knitr::opts_chunk$set(fig.width = 20, fig.height = 8);
require(ggplot2);
require(gridExtra);
```

## Synopsis

This report is trying to give give answer at two questions:  

* Across the United States, wich types of natural events are most harmful with respect to population health ?  
* Across the United States, which types of natural events have greatest economic consequences ?  
  
Storm Data dataset is an official publication of the National Oceanic and Atmospheric Administration (NOAA) which documents: 
 
* The occurrence of storms and other significant weather phenomena having sufficient intensity to cause loss of life, injuries, significant property damage, and/or disruption to commerce; 
 
* Rare, unusual, weather phenomena that generate media attention, such as snow flurries in South Florida or the San Diego coastal area; and 
 
* Other significant meteorological events, such as record maximum or minimum temperatures or precipitation that occur in connection with another event.  
 
  
## Data Analisys
### Data Processing, Loading and Processing the Raw Data
  
  
  
```{r dataset}

# if file doesn't exist, it should download it
if (!file.exists("StormData.csv.bz2")) {

	url = "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2";
	download.file(url, "StormData.csv.bz2");
	rm("url");
}

# read dataset into memory
dataset = read.csv("StormData.csv.bz2", na.string = ""
	, colClasses = c(rep("NULL", 7), "factor", rep("NULL", 14), rep("numeric", 3)
	, "character", "numeric", "character", rep("NULL", 9)));

# print summary
summary(dataset);

# print first 10 rows
head(dataset, 10);

```
  
  
  
#### Data Processing, Normalizing the dataset factors, K and/or k for thousends, M and/or m for millions, B and/or b for billions.
  
  
```{r}

bool = grepl("M|K|B", dataset$PROPDMGEXP, ignore.case = T);
dataset$PROPDMGEXP[!bool] = NA;
dataset$PROPDMGEXP = factor(toupper(dataset$PROPDMGEXP));

bool = grepl("M|K|B", dataset$CROPDMGEXP, ignore.case = T);
dataset$CROPDMGEXP[!bool] = NA;
dataset$CROPDMGEXP = factor(toupper(dataset$CROPDMGEXP));

```
```{r, echo=FALSE}
rm("bool");
```
  
  
  
#### Data Processing, Summarizing and subsetting the date for the first question, impact regarding the population health
```{r}

# calculate fatalities and injuries numbers by event
fatalities = tapply(dataset$FATALITIES, dataset$EVTYPE, sum, na.rm = T);
injuries = tapply(dataset$INJURIES, dataset$EVTYPE, sum, na.rm = T);

# order and chose first 10
f1 = as.data.frame(cbind(names(sort(fatalities, decreasing = T)[1:10])
		, unname(sort(fatalities, decreasing = T)[1:10])));
i1 = as.data.frame(cbind(names(sort(injuries, decreasing = T)[1:10])
		, unname(sort(injuries, decreasing = T)[1:10])));

names(i1) = c("NaturalEvent", "INJURIES");
names(f1) = c("NaturalEvent", "FATALITIES");

i1$INJURIES = as.integer(as.character(i1$INJURIES));
f1$FATALITIES = as.integer(as.character(f1$FATALITIES));

# print fatalities
print(f1);
# print injuries
print(i1);
```


#### Data Processing, Summarizing and subsetting the date for the second question, impact on economy
```{r}

# marmalizing the amounts of damages
# creating a new variable with normalized amount for property damage
dataset$PROPDMG_1 = dataset$PROPDMG;
# creating a new variable with normalized amount for crop damage
dataset$CROPDMG_1 = dataset$CROPDMG;

b = dataset$PROPDMGEXP == "K";
b[is.na(b)] = FALSE;
dataset$PROPDMG_1[b] = dataset$PROPDMG_1[b] * 1e+3;

b = dataset$PROPDMGEXP == "M";
b[is.na(b)] = FALSE;
dataset$PROPDMG_1[b] = dataset$PROPDMG_1[b] * 1e+6;

b = dataset$PROPDMGEXP == "B";
b[is.na(b)] = FALSE;
dataset$PROPDMG_1[b] = dataset$PROPDMG_1[b] * 1e+9;

b = dataset$CROPDMGEXP == "K";
b[is.na(b)] = FALSE;
dataset$CROPDMG_1[b] = dataset$CROPDMG_1[b] * 1e+3;

b = dataset$CROPDMGEXP == "M";
b[is.na(b)] = FALSE;
dataset$CROPDMG_1[b] = dataset$CROPDMG_1[b] * 1e+6;

b = dataset$CROPDMGEXP == "B";
b[is.na(b)] = FALSE;
dataset$CROPDMG_1[b] = dataset$CROPDMG_1[b] * 1e+9;

# creat a new variable with sum of property and crop damage
dataset$TOTALDMG = dataset$PROPDMG_1 + dataset$CROPDMG_1;

# compute total damage by event type
dmg = tapply(dataset$TOTALDMG, dataset$EVTYPE, sum, na.rm = T);
sort(dmg, decreasing = T)[1:10];

# order and chose first 10
d1 = as.data.frame(cbind(names(sort(dmg, decreasing = T)[1:10]), unname(sort(dmg, decreasing = T)[1:10])));

names(d1) = c("NaturalEvent", "TotalDamage");
# revert to "billions" as unit measuare
d1$TotalDamage = round(as.numeric(as.character(d1$TotalDamage)) * 1e-9, digits = 3);

print(d1);
```
```{r, echo=FALSE}
rm(list = c("fatalities", "injuries"));
```
## Results
### Plotting fatalities and injuries
  
  
  
```{r, fig.width=20, fig.height=16, fig.cap="Natural Events - impact on population health"}
# plot injuries
g1 = ggplot(i1, aes(x = NaturalEvent, y = INJURIES)) +
	geom_bar(stat = "identity", fill = "lightblue", color = "black") +
	theme_dark() + 
	ggtitle("Top recorded injuries across US caused by harmful natural event by type between 1950 and 2011") +
	geom_text(aes(label = INJURIES, vjust = -0.25)) +
	labs(y = "Recorded Injuries", x = "Natural Event Type");
# plot fatalities
g2 = ggplot(f1, aes(x = NaturalEvent, y = FATALITIES)) +
	geom_bar(stat = "identity", fill = "lightblue", color = "black") +
	theme_dark() + 
	ggtitle("Top recorded fatalities across US caused by harmful natural event by type between 1950 and 2011") +
	geom_text(aes(label = FATALITIES, vjust = -0.25)) +
	labs(y = "Recorded Fatalities", x = "Natural Event Type");

grid.arrange(g1, g2, nrow = 2);
```

### Plotting impact on economy

```{r, fig.width = 20, fig.height = 8, fig.cap="Natural Events - impact on economy"}
# plot damages
ggplot(d1, aes(x = NaturalEvent, y = TotalDamage)) +
	geom_bar(stat = "identity", fill = "lightblue", color = "black") +
	theme_dark() + 
	ggtitle("Top recorded damages across US caused by harmful natural event by type between 1950 and 2011") +
	geom_text(aes(label = paste('$', TotalDamage, " "), vjust = -0.25)) +
	labs(y = "Total Damage rounded in bilions, property and crops added", x = "Natural Event Type");

```
